### 2020-12-28 ch/2-related-work.tex
Add section to literature review that had been lost somehow

\chapter{Related work}\label{ch:related-work}
\section{Introduction}\label{introduction}
Fairness in Machine Learning is a topic that has recently received increasing attention.
Machine Learning has become very widespread in our society
which brings some dangers with it.
Decisions such as credit eligibility and hiring are increasingly done automatically.
However, the algorithms that are used in these decisions can have strong biases
that stem from the training data.
Often, the algorithms are even more biased than the training data.
\citet{kamishima2011fairness} give a great illustration for this
(adapted from \citet{calders2010three}).
The illustration is based on the Adult/Census Income dataset \citep{kohavi1996scaling}
which contains records of individuals with thirteen attributes, among them age and gender.
The task is to predict whether or not the individual earns more than \$50K per year.
In the dataset, 32\% of the men have a positive label (\ie, they earn more than \$50K),
but of the women only 11\% have a positive label.
With this ratio of about 3:1, the dataset is already skewed.
However, when training a \href{https://en.wikipedia.org/wiki/Naive_Bayes_classifier}{Naive Bayes classifier} on the data,
the ratio in the predictions is even more unfair:
the amount of women who get a positive prediction is even lower than 11\%.
If the Naive Bayes classifier does not get access to the gender information,
the predictions will be a bit fairer,
but they are still more unfair than the original dataset labels.

This phenomenon of predictions being more unfair than an already skewed dataset
is due to the generalization that classifiers have to make.
It is easier to classify all but the most outstanding women as low income
than to learn a complicated decision boundary.
Furthermore, removing the gender information does not solve the problem
because the gender can be inferred from other data.

The following is a review of the literature on Fairness in Machine Learning.
Several different approaches to and definition of Fairness are presented.
The focus is on identifying the major trends in the field from the beginning.

\subsection{Precursor}\label{precursor}
The first published work to address the problem of biases in machine learning
was arguably the 2008 paper by \citet{pedreshi2008discrimination}
with the main focus being Data Mining.
The first fundamental conclusion of the authors
is that it is not enough to take out the discriminatory features
(referred to as \emph{sensitive attributes} in other works) from the data.
That is, if the information about, for example, race
is removed from the data,
there is usually enough background information to reconstruct that attribute.
The experiments were done on the ``German credit dataset''
where the task is to classify people as a good or a bad credit risk.
As their focus is Data Mining,
the authors' goal is to identify discriminatory rules.
They distinguish between direct and indirect discrimination.
According to their definition, direct discrimination explicitly uses
discriminatory features in the premise of the rule, while indirect discrimination uses
features that are closely associated with discriminatory features. The latter case can be
detected by checking the other rules in the dataset.

\section{Modification of labels for Fairness enforcement}\label{modification-of-labels-for-fairness-enforcement}
Independently of \citet{pedreshi2008discrimination}, but with a similar intention,
\citet{kamiran2009classifying} considered discrimination in classification.
As opposed to \citet{pedreshi2008discrimination}, their goal was not to identify discriminatory rules in the dataset,
but to modify the dataset so that any classifier trained on it does not become biased.
The same dataset (German credit dataset) is used in both papers.

First, Kamiran and Calders present the problem formulation
that was later used in other works on Fairness as well:
there is a set of attributes, a binary class label
and a special \emph{sensitive attribute} which determines the demographic group.
The sensitive attribute \(s\) is assumed to be binary in the derivation,
but does not have to be.
\(s=0\) refers to the group vulnerable to discrimination and \(s=1\) to all other individuals.
It is further assumed that one of the class labels is generally desirable;
for example, it might correspond to being accepted for a credit or being given bail.
This class label is being referred to as the positive label: \(y = 1\).

@@

This measure of fairness is today called \emph{Demographic Parity}.
It is zero when the individuals in both groups have the same chance to get a positive prediction.

Their algorithm for de-biasing the dataset works as follows:
first, a well-calibrated classifier (Naive Bayes in the paper) is trained on the data
\emph{without} the sensitive attribute.
This classifier gives a ranking for how likely a positive label is for a given individual.
The highest ranked individuals that have label \(y=0\)
and are potentially discriminated against \(s=0\) get ``promoted'' to \(y=1\)
and the lowest ranked individuals with label \(y=1\) and \(s=1\) get ``demoted'' to \(y=0\)
until the dataset is fair according to the \(Disc\) measure.
According to the authors, the ranking ensures that the changes in the dataset
happen to the most appropriate candidates (close to the decision boundary),
the main goal being a prevention of a big drop in accuracy of the final classifier
with respect to the original dataset.

The paper concludes with results on the German credit data,
using Naive Bayes for ranking and also for the final classifier.
Using \emph{Age} as the sensitive attribute,
the dataset itself shows a \(Disc\) of about 15\% (depending on the train-test split).
As a baseline, a Naive Bayes classifier was trained with the sensitive attribute
resulting in \(Disc = 25\%\).
Without the sensitive attribute it leads to \(Disc = 10\%\).
The scheme that the authors propose reaches \(Disc = 5\%\).
The exact numbers depend on the split and the choice of sensitive attribute.
Generally, higher discrimination in the dataset itself
leads to bigger differences in discrimination between the fair classifier and the baseline.

This work was extended by \citet{calders2009building}.
They use the same discrimination criterion as before,
but call it \emph{dependency} between \(s\) and \(y\).
They again present the same dataset changing technique
and a different technique based on giving training examples different weights
instead of re-labeling them.
Examples with \(y=1\) and \(s=0\) get higher weight than those with \(y=0\) and \(s=0\).
For \(s=1\), those with \(y=0\) get higher weight than those with \(y=1\).
Weighting means that when the training examples are sampled from the dataset,
those with higher weight are used more often.
The two approaches are compared on the Adult/Census Income dataset \citep{kohavi1996scaling}
where the task is to estimate if someone earns more than \$50K per year.
The sensitive attribute is \emph{sex}.
In addition to Naive Bayes,
two classifiers based on nearest neighbors and a decision tree classifier are used.
The comparison shows that changing the labels leads to lower dependency (discrimination)
than reweighting the examples.
However, reweighting the examples leads to lower dependency than the baseline.
The difference in accuracy is minor.

\section{Fair classifiers}\label{fair-classifiers}
While the two previous papers were based on manipulating the dataset,
Calders and Verwer proposed three completely different approaches
to make classifiers fair in 2010 \citep{calders2010three}.
They all center around putting \emph{constraints} on the classifier make fair predictions.
This potentially gives more control over the prediction bias than manipulating datasets.
All three approaches are based on Naive Bayes
and again aim to enforce Demographic Parity as defined before.
The first approach shifts the predicted probabilities in a post-processing step
after training the classifier normally.
To this end, the sensitive attribute is treated differently
than the other features in the Naive Bayes model.
Instead of the class being the cause of the sensitive attribute,
the sensitive attribute is regarded as one cause for the class.
This is a significant change in perspective that enables a better intuition about the problem.
In this modified Naive Bayes,
the conditional probability \(P(y|s)\) is then modified in the already trained model
until the discrimination score is below a given threshold.
Care is taken to not distort the predicted distribution too much with respect to the unfair model.
The end result is a model that is (ideally) unbiased
but still relies on the sensitive attribute for predictions.
This is unworkable if the sensitive attributes are not always available when making predictions.
Nevertheless, the algorithm can be considered an improvement over manipulating the dataset,
as the resulting bias can be controlled to a much finer degree.

The second method that Calders and Verwer present in their 2010 paper
is based on training two separate models for each sensitive attributes.
That is, the dataset is split in two
where one half contains all examples with \(s=0\), we call this dataset \(D_0\),
and the other half contains all examples with \(s=1\), called \(D_1\).
For prediction, we choose either the model that was trained on \(D_0\) or the one for \(D_1\),
depending on the sensitive attribute of the example.
This method relies on the availability of the sensitive attribute for prediction,
like the previous method.
The two models are separately tweaked to produce overall fair results.
In order to minimize the effect on the accuracy, the models are tweaked the same amount each,
but in opposite directions.
Conceptually, having two different Naive Bayes model depending on the sensitive attribute
is equivalent to using one Naive Bayes model
in which the sensitive attribute is connected to all other features.
This means that the bias is in all of the features and not just in the class label.
The latter was the assumption of the previously discussed model.
The difference between these assumptions is not explored in much detail in the paper.

@@

In a comparison between \citet{kamishima2011fairness}'s algorithm
and the Naive Bayes method with two different models for the sensitive attributes
from \citet{calders2010three},
the Calders algorithm performed better on the Adult/Census Income dataset
(higher accuracy and lower discrimination score).
However, the Kamishima algorithm has the advantage
that the sensitive attribute does not necessarily have to be known for making predictions
and the trade-off between accuracy and fairness can be easily controlled
with the factor for the regularization term.

@@

The proposed idea is then as follows: a classifier is fair
if and only if the predictive distributions for any two data points
are at least as similar as the two points themselves,
according to a given similarity measure for distributions
and a given similarity measure for data points.
They call this condition the \emph{Lipschitz condition}.
The authors propose two practical similarity measures for distributions
that are well-known in the respective literature.
In order to train a fair classifier,
the Lipschitz condition is then used as a constraint for the optimization.
This is similar to how Kamishima et al.~used the fairness criterion as a regularizer.
However, Dwork et al.~intend to realize it as a strict constraint.
Furthermore, they show that their criterion is compatible with Demographic Parity
but does not guarantee it in general.
When the two demographic groups are similar according to the individual similarity measure,
then the Lipschitz condition will lead to a good approximation of Demographic Parity.
If the demographic groups are very different,
then the Lipschitz criterion is essentially incompatible with Demographic Parity.

@@

Finally, Dwork et al.~discuss the problem of removing sensitive information
from data that someone wants to sell.
The hope would be that the sold data has no information in it
that would allow a malicious actor to discriminate.
The authors argue that enforcing Demographic Parity would not be enough
to make the data unsuitable for discrimination.
They speculate that the Lipschitz condition would work better,
but ultimately do not provide a real solution to the problem.

\section{Fair representation}\label{fair-representation}
\citet{zemel2013learning} pick up on this idea of creating a \emph{fair representation}.
That is, the features of the data points are transformed such that they contain no bias anymore,
but still carry the information necessary for predicting the class label.
This is different from \citet{kamiran2009classifying} who also manipulated the dataset,
but only changed the \emph{labels}.
In this work, the \emph{inputs} are transformed.
This is the first of many papers with such a goal.
More specifically, the stated goal of \citet{zemel2013learning} is
to achieve both \emph{group fairness} (in the sense of Demographic Parity)
and \emph{individual fairness} (in the sense of treating similar people similarly),
as Dwork et al.~did before.
The learned fair representation is denoted by \(z\).
The condition for fairness is then
\begin{align}
  \label{eq:fair-representation}
  P(s=0|z=z') = P(s=1|z=z') ~.
\end{align}
Zemel et al.~propose to map the biased inputs \(x\) to \emph{prototypes} in the same space as \(x\).
The probability for being assigned to a particular prototype
must be the same for inputs with \(s=0\) and for \(s=1\).
Based on a given distance function (or similarity measure),
inputs are more likely to be assigned to prototypes that are close-by.
The classification task is then based entirely on the prototypes (the fair representation \(z\)).
In the paper, a linear model is used to map the prototypes to the outcomes \(y\).
Aside from the distance function, the whole model has therefore only two kinds of parameters:
the locations of the prototypes and the weights in the linear model.
These parameters are all optimized together.
The objective function consists of three terms:
the first enforcing Demographic Parity via the prototype locations,
the second ensuring that the prototypes are close to the inputs
and the third targeting a good accuracy by adapting the weights of the classifier.

The experiments are done on the German credit dataset, the Adult / Census Income dataset
and a dataset based on the Heritage Health Prize milestone 1 challenge
where the goal is to predict how many days a given person will spend in the hospital in a year.
The proposed methods performs better in these experiments
than the methods from Kamiran and Calders (2009) and the method from Kamishima.
However, they do not compare against Calders and Verwer (2010),
presumably because those methods rely on the sensitive attribute for making predictions,
which can be considered an unfair advantage.

A major flaw in Zemel et al.'s fair representation learning is
that the representation only enforces fairness
with respect to the one classifier that it was trained with.
This is shown in the paper when they try to predict \(s\) from \(z\) with a newly trained classifier
and get results that are significantly better than chance.
This shows that the proposed approach does not produce a representation
that could be sold to potentially malicious actors.
The other question is
whether the fair representation \(z\) still contains enough information about \(x\),
so that other classifiers can achieve high accuracy.
Here, the proposed system performs well:
a different classifier only shows a small reduction in accuracy
compared to the one that was trained together with the representation.

@@

For measuring the fairness of the (non-sensitive) input features
(disregarding class labels for the moment), the authors define \(\varepsilon\)-fairness.
The features \(x\) are \(\varepsilon\)-fair,
if any predictor that tries to predict \(s\) from \(x\) can only achieve a balanced error rate
that is higher than \(\varepsilon\).
Feldman et al.~prove that that \(\varepsilon\)-fairness with a suitable \(\varepsilon\)
is incompatible with violating the 80\% rule.
That is, if the sensitive attribute cannot be predicted from the features,
then a classifier trained on that data will automatically be fair (to a certain degree).
For datasets that are extremely unbalanced,
the required \(\varepsilon\) approaches 1/2
which corresponds to absolutely no information in \(x\) about \(s\).
Note that the definitions require that the performance of the best possible classifiers is known,
which is rarely the case.
In the paper, an SVM classifier is used to measure the \(\varepsilon\) for \(\varepsilon\)-fairness.

@@

In the experiments, the authors first investigate the theoretically predicted relationship
between \(\varepsilon\)-fairness and the \(DI\) measure.
All experiments were done with either of the two Partial Repair algorithms
which give very similar results.
Three datasets are used:
the German Credit dataset, the Adult/Census Income dataset
and the Ricci dataset from the \emph{Ricci vs DeStefano} court case.
The goal for the Ricci dataset is to predict which firefighters to promote based on test scores.
The sensitive attribute is \emph{race}.
The experimental results support the theoretical predictions to a reasonable extent.
To compare the method with previous work,
several classifiers are trained on the fair representation:
logistic regression, SVM and Naive Bayes.
Of these, the Naive Bayes classifier seems to outperform the algorithms from previous works
\citep{kamiran2009classifying,kamishima2011fairness,zemel2013learning}
in both achieved fairness and accuracy.

This work by \citet{feldman2015certifying} is a major improvement over the work by \citet{zemel2013learning}
because the representation (ideally) is fair with regards to any machine learning algorithm, not just one particular.
Furthermore, theoretical bounds were proved for the expected bias of a classifier.
However, Feldman et al.~do not take into account individual fairness in any way.

@@

This method can be improved further by taking into account the labels
when constructing the fair representation.
If this is not done, \(z\) loses the ranking information from \(x\) (see Feldman et al.~above).
To this end, a second latent variable is introduced, \(\tilde{z}\),
which encodes the variation in \(z\) that is not explained by the class labels \(y\).
\(z\) is then determined by \(y\) and \(\tilde{z}\),
and \(x\) is determined by \(z\) and \(s\) as before.
\(\tilde{z}\) and \(s\) have independent priors.
This structure ensures that \(y\) can be predicted from \(z\).
However, this introduces a new problem:
if \(y\) is correlated with \(s\), then \(z\) will be as well.
To overcome this, an additional penalty term is introduced that forces \(P(z|s=0)\)
and \(P(z|s=1)\) to be as close as possible.
This is realized with a measure of distance
between distributions called Maximum Mean Discrepancy (MMD).

The experiments were done on the Adult/Census Income dataset,
the German Credit dataset and the Heritage Health dataset.
To test whether \(s\) can be recovered from \(z\),
a Random Forest model and a Logistic Regression model were trained to predict \(s\) from \(z\).
This is compared to Zemel et al. (2013).
The proposed model can outperform Zemel et al.~on some tasks, but many results seem inconclusive.
Using MMD for an additional unfairness penalty, seems to improve fairness.
When training a classifier on \(z\) to predict \(y\), there is a small drop in accuracy.
In general, it is not clear whether this method gives better performance than Feldman et
al.~However, the algorithm by Louizos seems more general and more widely applicable
as it simply re-uses a widely used technique, variational autoencoders.

>>>>

\chapter{Related work}\label{ch:related-work}
%
% =========================
%      old introduction
% =========================
% \section{Introduction}\label{introduction}
% Fairness in Machine Learning is a topic that has recently received increasing attention.
% Machine Learning has become very widespread in our society
% which brings some dangers with it.
% Decisions such as credit eligibility and hiring are increasingly done automatically.
% However, the algorithms that are used in these decisions can have strong biases
% that stem from the training data.
% Often, the algorithms are even more biased than the training data.
% \citet{kamishima2011fairness} give a great illustration for this
% (adapted from \citet{calders2010three}).
% The illustration is based on the Adult/Census Income dataset \citep{kohavi1996scaling}
% which contains records of individuals with thirteen attributes, among them age and gender.
% The task is to predict whether or not the individual earns more than \$50K per year.
% In the dataset, 32\% of the men have a positive label (\ie, they earn more than \$50K),
% but of the women only 11\% have a positive label.
% With this ratio of about 3:1, the dataset is already skewed.
% However, when training a \href{https://en.wikipedia.org/wiki/Naive_Bayes_classifier}{Na\"ive Bayes classifier} on the data,
% the ratio in the predictions is even more unfair:
% the amount of women who get a positive prediction is even lower than 11\%.
% If the Na\"ive Bayes classifier does not get access to the gender information,
% the predictions will be a bit fairer,
% but they are still more unfair than the original dataset labels.
%
% This phenomenon of predictions being more unfair than an already skewed dataset
% is due to the generalization that classifiers have to make.
% It is easier to classify all but the most outstanding women as low income
% than to learn a complicated decision boundary.
% Furthermore, removing the gender information does not solve the problem
% because the gender can be inferred from other data.
%
% The following is a review of the literature on Fairness in Machine Learning.
% Several different approaches to and definition of Fairness are presented.
% The focus is on identifying the major trends in the field from the beginning.
\subsection{Two views of the dataset bias problem}\label{two-views-of-the-dataset-bias-problem}
As hinted to in the introduction, there is a divide in the related literature in how the problem is defined.
On the one hand, there is the ``groundtruth-centric'' view -- expressed through most of the introduction --
that the training set is biased but the deployment setting is not.
By \emph{deployment setting} I am here referring to the real distribution
that will be encountered by the machine learning system once it is used.

This can be seen as a form of transfer learning and notably does not exhibit a fairness-accuracy trade-off.

An instance of this setting has been formalized in \citet{blum2020recovering}.
In their formalization, the training set starts out unbiased,
but then, for samples with \(s=0\), class labels \(y\) are flipped from 1 to 0 with probability \(\nu\).
In addition, samples with \(s=0\) and \(y=0\) are dropped with probability \(1-\beta_\mathit{NEG}\)
and those with \(s=0\) and \(y=1\) are dropped with probability \(1-\beta_\mathit{POS}\).

On the other hand, there is the ``definition-centric'' view that given some training data,
and not necessarily knowing anything about the deployment setting,
we define fairness criteria which a fair classifier should satisfy.
Here, the goal is not necessarily to correct for a bias in the training data,
but simply to ensure that the classifier is fair in some specific sense.
By default, a classifier would \emph{not} be fair in that specific way.
It is left open whether this unfairness originates in the training data or in the algorithm itself.

This setting has a fairness-accuracy trade-off.

The definition-centric view was the one used by the first works in the field.
The first fairness definition was Demographic Parity. \textbf{citation}

The difference between these views is not completely clear-cut.
Consider the case where a biased training set is available,
and an unbiased test set is assumed to exist, but we do not have access to it.
Then we might at least know that the unbiased test set satisfies demographic parity.
In this case, testing the classifier for demographic parity is an indication
for how it would fare on the unbiased test set (which satisfies DP).

Furthermore, there is significant overlap in the solutions employed for the two views:
a method developed for one of them can often be used for the other as well.

The works discussed in the beginning of this chapter are mostly taking the definition-centric view.
Towards the end, I discuss more works with the other viewpoint.

\subsection{Precursor}\label{precursor}
The first published work to address the problem of biases in machine learning
was arguably the work by \citet{pedreshi2008discrimination}
with the main focus being data mining.
The first fundamental conclusion of the authors
is that it is not enough to take out the discriminatory features
(referred to as \emph{sensitive attributes} in other works) from the data.
That is, if the information about, for example, race
is removed from the data,
there is usually enough background information to reconstruct that attribute.
The experiments were done on the ``German credit dataset''
where the task is to classify people as a good or a bad credit risk.
In their context of data mining,
they distinguish between direct and indirect discrimination:
the former explicitly uses discriminatory features in the premise of the rule,
while the latter uses features that are closely associated with discriminatory features.
% The latter case can be
% detected by checking the other rules in the dataset.

\section{Modification of labels for Fairness enforcement}\label{modification-of-labels-for-fairness-enforcement}
Independently of \citet{pedreshi2008discrimination}, but with a similar intention,
\citet{kamiran2009classifying} considered discrimination in classification.
As opposed to \citet{pedreshi2008discrimination}, their goal was not to identify discriminatory rules in the dataset,
but to modify the dataset so that any classifier trained on it will be fair.
The same dataset (German credit dataset) is used in both papers.

The authors formulate the problem as it was later done in other works on fairness as well:
there is a set of features, a binary class label
and a special \emph{sensitive attribute} which determines the demographic group.
The sensitive attribute \(s\) is assumed to be binary in the derivation,
but does not have to be.
\(s=0\) refers to the group vulnerable to discrimination and \(s=1\) to all other individuals.
It is further assumed that one of the class labels is generally desirable;
for example, it might correspond to being accepted for a credit or being given bail.
This class label is being referred to as the positive label: \(y = 1\).

@@

This fairness metric corresponds to what is commonly referred to as \emph{demographic parity}.
It is zero when the individuals in both groups have the same chance to get a positive prediction.

Their algorithm for de-biasing the dataset works as follows:
first, a well-calibrated classifier (Na\"ive Bayes in the paper) is trained on the data
\emph{without} the sensitive attribute.
This classifier gives a ranking for how likely a positive label is for a given individual.
The highest ranked individuals that have label \(y=0\)
and are potentially discriminated against \(s=0\) get ``promoted'' to \(y=1\)
and the lowest ranked individuals with label \(y=1\) and \(s=1\) get ``demoted'' to \(y=0\)
until the dataset is fair according to the \(Disc\) measure.
According to the authors, the ranking ensures that the changes in the dataset
happen to the most appropriate candidates (close to the decision boundary),
the main goal being a prevention of a big drop in accuracy of the final classifier
with respect to the original dataset.

The paper concludes with results on the German credit data,
using Na\"ive Bayes for ranking and also for the final classifier.
Using \emph{Age} as the sensitive attribute,
the dataset itself shows a \(Disc\) of about 15\% (depending on the train-test split).
As a baseline, a Na\"ive Bayes classifier was trained with the sensitive attribute
resulting in \(Disc = 25\%\).
Without the sensitive attribute it leads to \(Disc = 10\%\).
The scheme that the authors propose reaches \(Disc = 5\%\).
The exact numbers depend on the split and the choice of sensitive attribute.
Generally, higher discrimination in the dataset itself
leads to bigger differences in discrimination between the fair classifier and the baseline.

This work was extended in \citet{calders2009building},
where the same discrimination criterion is used as before,
but called \emph{dependency} between \(s\) and \(y\).
They again present the same dataset changing technique
and a different technique based on giving training examples different weights
instead of re-labeling them.
Examples with \(y=1\) and \(s=0\) get higher weight than those with \(y=0\) and \(s=0\).
For \(s=1\), those with \(y=0\) get higher weight than those with \(y=1\).
Weighting means that when the training examples are sampled from the dataset,
those with higher weight are used more often.
The two approaches are compared on the Adult/Census Income dataset \citep{kohavi1996scaling}
where the task is to estimate if someone earns more than \$50K per year.
The sensitive attribute is \emph{sex}.
In addition to Na\"ive Bayes,
two classifiers based on nearest neighbors and a decision tree classifier are used.
The comparison shows that changing the labels leads to lower dependency (discrimination)
than reweighting the examples.
However, reweighting the examples leads to lower dependency than the baseline.
The difference in accuracy is minor.

\section{Fair classifiers}\label{fair-classifiers}
Another common approach is to put fairness \emph{constraints} on the classifier
instead of manipulating the dataset.
\citet{calders2010three} proposes three such approaches,
which potentially give more control over the prediction bias than changing the datasets.
All three approaches are based on Na\"ive Bayes
and again aim to enforce Demographic Parity in the predictions as defined before.
The first approach shifts the predicted probabilities in a post-processing step
after training the classifier normally.
To this end, the sensitive attribute is treated differently
than the other features in the Na\"ive Bayes model.
Instead of the class being the cause of the sensitive attribute,
the sensitive attribute is regarded as one cause for the class.
This is a significant change in perspective that enables a better intuition about the problem.
In this modified Na\"ive Bayes,
the conditional probability \(P(y|s)\) is then modified in the already trained model
until the discrimination score is below a given threshold.
Care is taken to not distort the predicted distribution too much with respect to the unfair model.
The end result is a model that is (ideally) unbiased
but still relies on the sensitive attribute for predictions.
This is unworkable if the sensitive attributes are not always available when making predictions.
Nevertheless, the algorithm can be considered an improvement over manipulating the dataset,
as the resulting bias can be controlled to a much finer degree.

The second method in \citet{calders2010three}
is based on training two separate models for each sensitive attributes.
That is, the dataset is split in two
where one half contains all examples with \(s=0\), we call this dataset \(D_0\),
and the other half contains all examples with \(s=1\), called \(D_1\).
For prediction, we choose either the model that was trained on \(D_0\) or the one for \(D_1\),
depending on the sensitive attribute of the example.
This method relies on the availability of the sensitive attribute for prediction,
like the previous method.
The two models are separately tweaked to produce overall fair results.
In order to minimize the effect on the accuracy, the models are tweaked the same amount each,
but in opposite directions.
Conceptually, having two different Na\"ive Bayes model depending on the (binary) sensitive attribute
is equivalent to using one Na\"ive Bayes model
in which the sensitive attribute is connected to all other features.
This means that the bias is in all of the features and not just in the class label.
The latter was the assumption of the previously discussed model.
The difference between these assumptions is not explored in much detail in the paper.

@@

In a comparison between \citet{kamishima2011fairness}'s algorithm
and the Na\"ive Bayes method with two different models for the sensitive attributes from \citet{calders2010three},
the latter performed better on the Adult/Census Income dataset
(higher accuracy and lower discrimination score).
However, the former has the advantage
that the sensitive attribute does not necessarily have to be known for making predictions
and the trade-off between accuracy and fairness can be easily controlled
with the factor for the regularization term.

@@

The proposed idea is then as follows: a classifier is fair
if and only if the predictive distributions for any two data points
are at least as similar as the two points themselves,
according to a given similarity measure for distributions
and a given similarity measure for data points.
They call this condition the \emph{Lipschitz condition}.
The authors propose two practical similarity measures for distributions
that are well-known in the respective literature.
In order to train a fair classifier,
the Lipschitz condition is then used as a constraint for the optimization.
This is similar to how \citet{kamishima2011fairness} used the fairness criterion as a regularizer.
However, \citet{dwork2012fairness} intend to realize it as a strict constraint.
Furthermore, they show that their criterion is compatible with Demographic Parity
but does not guarantee it in general.
When the two demographic groups are similar according to the individual similarity measure,
then the Lipschitz condition will lead to a good approximation of Demographic Parity.
If the demographic groups are very different,
then the Lipschitz criterion is essentially incompatible with Demographic Parity.

@@

Finally, \citet{dwork2012fairness} discuss the problem of removing sensitive information
from data that someone wants to sell.
The hope would be that the sold data has no information in it
that would allow a malicious actor to discriminate.
The authors argue that enforcing Demographic Parity would not be enough
to make the data unsuitable for discrimination.
They speculate that the Lipschitz condition would work better,
but ultimately do not provide a real solution to the problem.

\section{Fair representation}\label{fair-representation}
\citet{zemel2013learning} expand on the idea of creating a \emph{fair representation}.
That is, the features of the data points are transformed such that all bias is removed
but still carry the information necessary for predicting the class label.
This is different from \citet{kamiran2009classifying} where the dataset was manipulated as well,
but the change was restricted to the \emph{labels}.
In this work, the \emph{inputs} are transformed.
This is the first of many works with such a goal.
More specifically, the stated goal of \citet{zemel2013learning} is
to achieve both \emph{group fairness} (in the sense of Demographic Parity)
and \emph{individual fairness} (in the sense of treating similar people similarly),
as \citet{dwork2012fairness} did before.
The learned fair representation is denoted by \(z\).
The condition for fairness is then
\begin{align}
  \label{eq:fair-representation}
  P(s=0|z=z') = P(s=1|z=z') ~.
\end{align}
\citet{zemel2013learning} propose to map the biased inputs \(x\) to \emph{prototypes} in the same space as \(x\).
The probability for being assigned to a particular prototype
must be the same for inputs with \(s=0\) and for \(s=1\).
Based on a given distance function (or similarity measure),
inputs are more likely to be assigned to prototypes that are close-by.
The classification task is then based entirely on the prototypes (the fair representation \(z\)).
In the paper, a linear model is used to map the prototypes to the outcomes \(y\).
Aside from the distance function, the whole model has therefore only two kinds of parameters:
the locations of the prototypes and the weights in the linear model.
These parameters are all optimized together.
The objective function consists of three terms:
the first enforcing Demographic Parity via the prototype locations,
the second ensuring that the prototypes are close to the inputs
and the third targeting a good accuracy by adapting the weights of the classifier.

The experiments are done on the German credit dataset, the Adult / Census Income dataset
and a dataset based on the Heritage Health Prize milestone 1 challenge
where the goal is to predict how many days a given person will spend in the hospital in a year.
The proposed methods performs better in these experiments
than the methods from \citet{kamiran2009classifying} and the method from \citet{kamishima2011fairness}.
However, they do not compare against \citet{calders2010three},
presumably because those methods rely on the sensitive attribute for making predictions,
which can be considered an unfair advantage.

A major flaw in \citet{zemel2013learning}'s fair representation learning is
that the representation only enforces fairness
with respect to the one classifier that it was trained with.
This is shown in the paper when they try to predict \(s\) from \(z\) with a newly trained classifier
and get results that are significantly better than chance.
This shows that the proposed approach does not produce a representation
that could be sold to potentially malicious actors.
The other question is
whether the fair representation \(z\) still contains enough information about \(x\),
so that other classifiers can achieve high accuracy.
Here, the proposed system performs well:
a different classifier only shows a small reduction in accuracy
compared to the one that was trained together with the representation.

@@

For measuring the fairness of the (non-sensitive) input features
(disregarding class labels for the moment), the authors define \(\varepsilon\)-fairness.
The features \(x\) are \(\varepsilon\)-fair,
if any predictor that tries to predict \(s\) from \(x\) can only achieve a balanced error rate
that is higher than \(\varepsilon\).
\citet{feldman2015certifying} prove that that \(\varepsilon\)-fairness with a suitable \(\varepsilon\)
is incompatible with violating the 80\% rule.
That is, if the sensitive attribute cannot be predicted from the features,
then a classifier trained on that data will automatically be fair (to a certain degree).
For datasets that are extremely unbalanced,
the required \(\varepsilon\) approaches 1/2
which corresponds to absolutely no information in \(x\) about \(s\).
Note that the definitions require that the performance of the best possible classifiers is known,
which is rarely the case.
In the paper, an \ac{SVM} classifier is used to measure the \(\varepsilon\) for \(\varepsilon\)-fairness.

@@

In the experiments, the authors first investigate the theoretically predicted relationship
between \(\varepsilon\)-fairness and the \(DI\) measure.
All experiments were done with either of the two Partial Repair algorithms
which give very similar results.
Three datasets are used:
the German Credit dataset, the Adult/Census Income dataset
and the Ricci dataset from the \emph{Ricci vs DeStefano} court case.
The goal for the Ricci dataset is to predict which firefighters to promote based on test scores.
The sensitive attribute is \emph{race}.
The experimental results support the theoretical predictions to a reasonable extent.
To compare the method with previous work,
several classifiers are trained on the fair representation:
logistic regression, SVM and Na\"ive Bayes.
Of these, the Na\"ive Bayes classifier seems to outperform the algorithms from previous works
\citep{kamiran2009classifying,kamishima2011fairness,zemel2013learning}
in both achieved fairness and accuracy.

This work by \citet{feldman2015certifying} is a major improvement over the work by \citet{zemel2013learning}
because the representation (ideally) is fair with regards to any machine learning algorithm, not just one particular.
Furthermore, theoretical bounds were proved for the expected bias of a classifier.
However, the work does not take into account individual fairness in any way.

@@

This method can be improved further by taking into account the labels
when constructing the fair representation.
If this is not done, \(z\) loses the ranking information from \(x\) (see \citet{feldman2015certifying} above).
To this end, a second latent variable is introduced, \(\tilde{z}\),
which encodes the variation in \(z\) that is not explained by the class labels \(y\).
\(z\) is then determined by \(y\) and \(\tilde{z}\),
and \(x\) is determined by \(z\) and \(s\) as before.
\(\tilde{z}\) and \(s\) have independent priors.
This structure ensures that \(y\) can be predicted from \(z\).
However, this introduces a new problem:
if \(y\) is correlated with \(s\), then \(z\) will be as well.
To overcome this, an additional penalty term is introduced that forces \(P(z|s=0)\)
and \(P(z|s=1)\) to be as close as possible.
This is realized with a measure of distance
between distributions called Maximum Mean Discrepancy (MMD).

The experiments were done on the Adult/Census Income dataset,
the German Credit dataset and the Heritage Health dataset.
To test whether \(s\) can be recovered from \(z\),
a Random Forest model and a Logistic Regression model were trained to predict \(s\) from \(z\).
This is compared to \citet{zemel2013learning}.
The proposed model can outperform \citet{zemel2013learning} on some tasks, but many results seem inconclusive.
Using MMD for an additional unfairness penalty, seems to improve fairness.
When training a classifier on \(z\) to predict \(y\), there is a small drop in accuracy.
In general, it is not clear whether this method gives better performance than Feldman et
al.~However, the algorithm by Louizos seems more general and more widely applicable
as it simply re-uses a widely used technique, variational autoencoders.

\section{Other fairness criteria}%
\label{improved-definitions-of-fairness}
The early literature on fairness in machine learning used only \ac{DP}
and similarity-based fairness to judge the fairness of a classifier.
\citet{kleinberg2016inherent} formalized three different fairness conditions.
The authors consider a scenario where the data points \(x\) are sorted into bins \(b\)
and each bin is associated with a prediction score \(f = P(\hat{y}=1|b)\)
where \(\hat{y}\) is the predicted class label and \(x\) the (non-sensitive) features.
\begin{enumerate}
\item
  Calibration within groups:
  If the prediction score for a given bin is \(f\),
  then when considering all the data points with group \(s\) in the bin,
  a fraction of \(f\) of those should have the class label \(y=1\).
  In other words, the prediction score \(f\) is well calibrated with respect to group \(s\).
  This should be the case for all groups.
\item
  Balance for the negative class:
  The true negative rate should be the same for both groups: \(P(\hat{y}=0|y=0,s=0) = P(\hat{y}=0|y=0,s=1)\).
\item
  Balance for the positive class:
  The true positive rate should be the same for both groups: \(P(\hat{y}=1|y=1,s=0) = P(\hat{y}=1|y=1,s=1)\).
\end{enumerate}
These are formalizations of things that have been proposed as definitions of fairness.
The first criterion essentially ensures that the predictor works correctly for both groups.
This prevents a classifier from predicting one group correctly
but always returning a negative answer for the other group.
The second and third put emphasis on negative and positive class labels respectively.
In criterion 2, we allow the classifier to mis-classify those data points with \(y=0\),
but we want to make sure that the misclassification rate is the same for both groups.
In other words, members of the different groups have the same chance
to get a correct classification if they should receive a negative classification.
The same holds for criterion 3 and positive classifications (\(y=1\)).
